\begin{acknowledgements}
\addchaptertocentry{\acknowledgementname} % Add the acknowledgements to the table of contents

Research, the verb, is as any other skill, a thing that you develop only with time and
practice. I would lie if I did not say that the last two years have felt like an intensive
camp to learn to research. Research is hard, but it is harder to try to explain what you
have done to others, but as I discovered writing this document and taking notes, writing
what you think it is one of the most useful and important parts of the \textit{journey}.
Research is not always fun, failing at it is depressing, and as if it was not enough,
research is often boring and arduous. Fortunately, I was not alone in this journey, I have
been surrounded by masters in this art and I would like to thank them.

To my advisors:

\begin{itemize}
\tightlist
\item Felipe Restrepo, who has been very supportive and who has held wave after wave of my
  bad written texts.
\item Fabio González, who has been flexible enough to let me explore different areas on my
  own. Without his critics on my wild ideas, I would not have put the effort to show that
  some of those ideas actually had some substance.
\end{itemize}

To the staff and people at the university:

\begin{itemize}
\tightlist
\item Jonatan Goméz, who told me with years of anticipation what I should not be doing.
  His knowledge in what could be wrong has always been an inspiration to do better.
\item Rodrigo Moreno, Oscar Perdomo, and all the others in the Mindlab group for their
  support and company in the long and arduous days at the university.
\item To all my students, who showed me how hard it is to actually program human minds.
  Some, very susceptible to change and others impossible to do so.
\end{itemize}

No human being can live without others. We all need a social net, something to support
ourselves. I would like to acknowledge their time on keeping me distracted from insanity.

To my friends:

\begin{itemize}
\tightlist
\item Sergio Dorado, who motivated me to do my best and find new goals. He infected me--I
  envy him--with his ideas and accomplishments.
\item Helen Smith, who, even 8 thousand kilometres away, was able to find time to chat
  with this Colombian guy.
\item William, Solene and Vinicius for their time and memories.
\item Vicky, whose perseverance I envy, and whose work and learning ethic I try to imitate.
\end{itemize}

To the external events that delayed or accelerated this thesis culmination:

\begin{itemize}
\tightlist
\item To the \enquote{paro,} a students` university strike, for giving me 3 extra months to
  write this document,  and
\item To the sicknesses of my family members, which helped me centre myself in what is
  important.
\end{itemize}

I could go on and on. There are hundreds of people and little pieces of
reality that made me who I am, and who helped me to get to the end. I am
grateful of having them all, albeit my faulty memory is not able to remember them all.
Sorry you guys.

But certainly, I could have never done a thing without the help of my
family:

\begin{itemize}
\tightlist
\item Isabel Camacho, my mom, who has done all the boring tasks to let me concentrate
  on the work, I owe her my life (and not only literally).
\item María del Carmen Gordillo, my grandma, who has always been very attentive to my and
  my mom's health.
\item My cousins Paola, Laura, Danilo, Angie, Duvan, and Karol who have always given me
  some excuse to relax, and from whom I have learnt so much about the world.
\item My aunties Martha, Teresa and Patricia, and my Uncle Juan for their support.
\item My dad, without whom I would not be here.
\end{itemize}

To all, thank you for being there.

\end{acknowledgements}
