\chapter{Introduction}\label{introduction}

\section{Motivation}\label{motivation}

Dynamically typed programming languages (Dynamically typed languages for short) like
Python and Ruby have risen in popularity over the last two decades. Their simple syntax,
ease of use, out-of-the-box REPLs\footnote{Read-eval-print loop}, dynamic type systems,
and their great number of libraries have been the main factors to this rise. Indeed,
developers can to write, execute and see results immediately in fast development cycles,
especially at the initial stages. However, their simple syntax and dynamic typing makes
them particularly difficult to analyse as the more rigid the language, the easier to
analyse it is. Most efforts on Static Analysis for Python have focused on assuming some
restrictions on the language. For example, MyPy \autocite{lehtosalo2016mypy} assumes the
type of variables as static.

Dynamically typed languages only check type mismatches at runtime, i.e.~errors between
variable types only get detected when the code is run. For example,
consider the following piece of Python code:

\begin{pythoncode*}{gobble=2}
  myvar = "Some text in a string"
  other = 4.1
  result = "oh my" + 2  # This will fail
  print(result + other, myvar)
\end{pythoncode*}

The code is syntactically correct, but it fails to execute. Python will interpret the first
two lines and will stop in the third, where it will throw an exception because strings
and numbers cannot be added.

A Static Type Analysis tool, such as MyPy \autocite{lehtosalo2016mypy}, is able to detect
type mismatches without ever executing the code.  MyPy has the ability to infer the type
of variables, e.g.~from the previous example, it detects that \pycode/other/ is of type
\pycode/float/ and \pycode/myvar/ of type \pycode/str/.

Python dynamic nature makes Python extraordinarily hard to statically analyse. In fact,
due to Python introspection capabilities, it is impossible to warranty the behaviour of a
piece of Python code without assuming some fixed basic semantics in the interpreter.  MyPy
tackles the undecidability problem of Python dynamic nature by restricting the number
of valid Python programs to those that conform to a static type system\footnotemark.

\footnotetext{I would like to express my inmense gratitude to the MyPy team for writing
  such an amazing tool. I do not know how many hours they saved me from endless
  frustration at debugging. What I like more about MyPy is that writing code with it feels
  just remarkably similar to writing code in language with a stronger static type system
  (e.g.~Haskell).}

Even with some assumptions in place, the inference ability of MyPy infers the type of
variables, but not the type of functions. MyPy allows optionally to add type annotations
to the code, to either variables or functions. Type annotations allow more restrictive and
precise types, and help MyPy to catch more potential bugs.

Although MyPy is capable of checking the correctness of code, its assumptions limit it from a whole class of valid and correct
Python code\footnotemark. There are many scenarios where even with type annotations, MyPy
cannot detect a bug in the code. For example, consider the following piece of python
code:%\footnotemark:

\footnotetext{For example, there is no way (currently) to make MyPy happy about the
  function \pycode/def addtwice(x, y): return x + y + y; addtwice(3,2); addtwice('a','b')/.
  If no type annotations is given, MyPy alerts the user on the usage of non-typed
  functions. But no type can satisfy the assertion because of the \enquote{polymorphisity}
  of the \pycode/+/ operation.}

%\footnotetext{Notice the lack of type annotations in this example. MyPy is able to infer the
%  type of the variables in this example.}

\begin{pythoncode*}{gobble=2}
  import numpy as np
  x = np.array( [[1,2,3], [4,5,6]] ) # shape (2,3)
  y = np.array( [[7], [0], [2], [1]] ) # shape (4,1)
  z = np.dot( x, y ) # trying to apply dot product
\end{pythoncode*}

In this example, we are using NumPy \autocite{oliphant2006guide}, which is a well known
Python library, but it could also have been TensorFlow \autocite{abadi_tensorflow_2016} or
any other library with support for tensors and tensor operators. Tensors are a
generalisation of vectors and arrays.  An array of size \texttt{n}, or a tensor with shape
\texttt{(n,)}, contains exactly \(n\) elements. A matrix of size \(n \times m\),
alternatively, a tensor with shape \texttt{(n,\ m)}, is a structure that holds
\(n \cdot m\) elements. Notice that a shape is any tuple of natural number and a tensor is
a structure with a shape. For example, the shape \texttt{(2,\ 5,\ 6)} indicates us a
tensor with \(2 \cdot 5 \cdot 6 = 60\) elements.

The example above will fail to run as it hits an erroneous condition at the last line. Two
arrays/tensors are created, \texttt{x} and \texttt{y} with shapes \texttt{(2,\ 3)} and
\texttt{(4,\ 1)}, respectively. Then, a matrix multiplication with both tensors tries to
be performed, but it fails because the shapes of the two matrices are incompatible
(\(3\ne4\)). The call to \texttt{np.dot} fails, but the developer will not be warned about
it and will only notice it when executing the code. To summarise, the aforementioned piece
of code type checks in MyPy even though it fails to run.

If it were possible to add the shape of the tensors as part of the type definitions, we
could potentially type check for mismatches of tensor shapes. Unfortunately, there does
not seem to be a straightforward way to add the shape of tensors to their types and check
them in MyPy.

Tensors allow writing computations more concisely. Tensor operations are also highly
optimisable and can be run in parallel. With the advent of Deep Learning, tensors have
become a central figure in developers toolkit.  Tensor operations fail if the tensors are
not of the right shape. To statically analyse if operations between tensors are valid, the
shape of tensors must be computed. To compute the shape of tensors, we require to compute
also the value of every other variable in the language. The type of static analysis that
is focused on finding out the value of variables is called Static Value Analysis.

\section{Problem Definition}\label{problem-definition}

Tensors are becoming an essential abstraction in the toolkit of
developers. Tools tailored to work with or based on tensors have been
popping out in recent years. It is therefore imperative to develop
analysis tools to verify and flag potential bugs. A Static Value
Analysis tool focused on tensor shape analysis could aid developers in
their work, as it could be able to flag potential bugs related to the
shapes of tensors as the developer writes code.

All of the theoretical approaches that tackle the problem focus on type
checking tensors and tensor operations. \textcite{griffioen_type_2015},
\textcite{slepak_array-oriented_2014}, and \textcite{rink_modeling_2018}
define a type system extended with tensors and tensor operations. The
principal idea of these type systems is to encode the restrictions that
the operations on tensors require, a tensor operation is valid if the
restriction checks with the tensors.

Practical attempts have been mainly focused on languages with strong
type systems, and recently some in Python
\autocites{fromherz_static_2018}{monat_static_2018}.
\textcite{chen_typesafe_2017} (in Scala) and
\textcite{eaton_statically_2006} (in Haskell) have tried to annotate
tensors' types with their shapes, leaving the compiler's type check
inference system to check the shapes. At the time
\textcite{eaton_statically_2006} proposed his methodology to extend
types with additional data, GHC (Glasgow Haskell Compiler), the by
default Haskell compiler, didn't have many capabilities to work with
data at the type level, resulting in some rather complicated code to
read and write. A recent approach to type check tensors in Haskell has
been shown in a library written by \textcite{elkin_haskell_2018}, which
uses the updated Dependent Type System of Haskell to code and enforce
type shapes.

Abstract Interpretation is a framework for static analysis. The main
idea of Abstract Interpretation is to overapproximate the result of
computing a piece of code. Any piece of code can be run in an Abstract
Interpreter, an interpreter build from the semantics defined by Abstract
Interpretation, in a finite amount of time. To warranty that the
Abstract Interpreter runs in a finite amount of time, Abstract
Interpretation defines a series of rules that force to overapproximate
the result of executing operations by the program. Abstract
Interpretation is especially useful for Static Value Analysis as the
rules to derive the semantics of an Abstract Interpreter from the formal
semantics of the language are very well studied, formalised and explicit
\autocite{cousot_abstract_1977}.

Writing an Abstract Interpreter for Python requires the formal semantics
of Python as a starting point. Unfortunately, Python does not have any
official formal semantics defined. Some attempts to define a formal
semantics for Python have been done
\autocites{politz_python_2013}{fromherz_static_2018}{guth_formal_2013}{ranson_semantics_2008}
but none of them take into account type annotations.

\textbf{Problem:} Building an Abstract Interpreter for Python to
Statically Value-Analyse tensor operations.

{\inlinetodo{Should I add some research questions?}}

\subsubsection{Assumptions and Scope}\label{assumptions-and-scope}

Very little can be asserted from a piece of code in a programming
language like Python without making any assumptions about the
environment. For example, consider the following piece of code:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# runme.py}
\NormalTok{a }\OperatorTok{=} \BuiltInTok{int}\NormalTok{(}\StringTok{'6'}\NormalTok{) }\OperatorTok{+} \DecValTok{2}
\BuiltInTok{print}\NormalTok{(}\DecValTok{7}\NormalTok{)}
\ControlFlowTok{assert}\NormalTok{ a }\OperatorTok{==} \DecValTok{8}
\end{Highlighting}
\end{Shaded}

One may be tempted to say that the code above never fails, but it would
be mistaken. Thanks to Python's introspection capabilities, the code
above could do just about anything. A piece of Python code can be run
directly in the interpreter or executed by an \VERB|\BuiltInTok{exec}|
or \VERB|\BuiltInTok{eval}| statement. The functions \texttt{int} and
\texttt{print} can be redefined to compute anything we want. For
example, consider the following piece of code in that calls
\texttt{runme.py}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{runme }\OperatorTok{=} \BuiltInTok{open}\NormalTok{(}\StringTok{'runme.py'}\NormalTok{).read()}
\NormalTok{fakeinit }\OperatorTok{=} \StringTok{"""}
\StringTok{def int(val):}
\StringTok{  return 3.0}
\StringTok{def print(val):}
\StringTok{  global a}
\StringTok{  a = val}
\StringTok{"""}
\BuiltInTok{exec}\NormalTok{(fakeinit }\OperatorTok{+}\NormalTok{ runme, \{\})  }\CommentTok{# Throws an exception}
\end{Highlighting}
\end{Shaded}

From now on, we will assume that any piece of code written in Python
will be called from the interpreter directly without changing the
behaviour of any builtin function or variable (this includes the
behaviour of variables and functions from non-builtin libraries as
NumPy), i.e.~we assume that the code will never be run by an
\VERB|\BuiltInTok{exec}| call but always \texttt{import}ed or
interpreted directly by \texttt{python\ file-code.py}.

As a developer, I can say that I like when I see a piece of code flag
wrong and IT IS really WRONG, but I despise when the opposite happens.
It is very annoying to have a tool flag every sentence you write as
wrong even though the code works just fine. We will assume from here
onwards that we want the Tensor Analysis to flag only errors that are
going to happen if the code is run, i.e.~we want only true positives not
maybe positives (as a lack of a better word).

Although, TensorFlow \autocite{abadi_tensorflow_2016}, PyTorch
\autocite{paszke2017pytorch} and other libraries would benefit from the
development of a tool targeted to check them, I chose to check NumPy's
tensor operations. NumPy is the standard library for computing with
tensors and it is used as backend on most projects.

Implementing an Abstract Interpreter, or any Static Analysis, for a
language like Python is a considerable undertaking, mainly because of
the breadth of characteristics a mature Programming Language like Python
has. Thus, we will centre on just a couple of Python core
characteristics and will leave others for future work. The
characteristics explored in the current document are:

\begin{itemize}
\tightlist
\item
  Builtin primitive variables: \texttt{int}, \texttt{float},
  \texttt{bool}, \texttt{None}, \texttt{list}, \texttt{tuple}
\item
  Primitive functions: \texttt{int}, \texttt{print}, \texttt{input},
  \ldots{}
\item
  Boolean and numeric operations: \texttt{+}, \texttt{-}, \texttt{*},
  \texttt{/}, \texttt{\%}, \texttt{**}, \texttt{\textless{}\textless{}},
  \texttt{\textgreater{}\textgreater{}}, \texttt{//},
  \texttt{\textless{}}, \texttt{\textless{}=}, \texttt{\textgreater{}}
  and \texttt{\textgreater{}=}
\item
  \texttt{if} and \texttt{while} staments
\item
  Import statement (limited only to the \texttt{numpy} library)
\item
  NumPy arrays and some of its methods/functions to work with NumPy
  arrays (including \texttt{dot}, \texttt{zeros}, \texttt{shape}, and
  numeric operations with broadcasting)
\end{itemize}

Consider the following piece of code:

\begin{verbatim}
if someval:
  i = 3
else:
  i = "ntr"
print(i)
\end{verbatim}

The variable \texttt{i} can be either an \texttt{int} or a \texttt{str},
thus the type of \texttt{i} should be \texttt{Union{[}int,str{]}}. If
the type of a variable is an \texttt{Union} type then we need to build
an Abstract Interpreter (actually, an Abstract Domain) aware of unions.
Building an Abstract Interpreter aware of \texttt{Union} types is not
considered part of this work, therefore if a variable holds more than
one value, e.g.~it is either a \texttt{5} or \texttt{2.3}, then the
variable's type will be \texttt{Any} and not \texttt{Union}.
\texttt{Any} and \texttt{Union} types are defined in the \texttt{typing}
library \autocite{pep484}, an implementation of Gradual Typing\footnote{Gradual
  Typing is a kind of Static Type Analysis targeted to Dynamically Typed
  Languages like Python}.

Gradual Typing restricts the number of valid programs to those that type
check. A variable in Gradual Typing is of a specific type and it never
changes, i.e.~if a variable's type is set to be \texttt{int} then it can
only contain \texttt{int}s. \texttt{Union} types are meant to indicate
that a variable may take more than one value at any point in time, which
would mean that any variable with a \texttt{Union} type would
automatically become an \texttt{Any} in this work, but that is not the
truth. Consider the following piece of code:

\begin{verbatim}
i = 3
i = "ntr"
i += "ueor"
print(i)  # the type of `i` is `str`
\end{verbatim}

Under Gradual Typing, the type of \texttt{i} is
\texttt{Union{[}int,str{]}}, but we know that the variable \texttt{i}
holds only one type at \textbf{all} times, i.e. \texttt{i} never holds
two or more values as in the previous example. An Abstract Interpreter
would run each line sequentially, and it would never find a that the
value of \texttt{i} is both \texttt{int} and \texttt{str} at the same
time, i.e.~the variable is never considered to be of type \texttt{Any}
but it will have type \texttt{int} and then type \texttt{str} as the
code is run. In this regard, an Abstract Interpreter can give a better
approximation of what the types of a variable are at any point in time,
while Gradual Typing considers all possible types a variable may have at
any point in time.

To recapitulate, the assumptions and scope of the thesis are:

\begin{itemize}
\tightlist
\item
  The behaviour of builtins and imported variables and functions is
  always the same and is determined by the Python reference manual or
  the library's author, e.g.~\texttt{input} is a function that returns
  a \texttt{str}.
\item
  The user will only be reported errors that will happen but no errors
  that \emph{may} happen (This means, the resulting tool is not a
  verification tool).
\item
  The Static Value Analysis centres around checking operations from the
  NumPy library.
\item
  Only a selected subset of Python characteristics and functions are
  explored.
\item
  A variable can hold only one type of value at the time. If a variable
  is set to hold values of different types at the same time, then its
  value will be \texttt{Any}.
\end{itemize}

\section{Objectives}\label{objectives}

\subsection{General objective}\label{general-objective}

To design and implement a strategy for statically analyse tensor shapes
in Python to support early detection of potential bugs before execution.

\subsection{Specific objectives}\label{specific-objectives}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To design and implement an abstract interpreter for Python.
\item
  To design and implement a strategy that uses the abstract interpreter
  to analyse the shapes of tensors for Python code.
\item
  To implement a tool that flags the bugs inferred by the abstract
  interprete.
\item
  To empirically evaluate the developed static analysis in a set of
  representative test cases.
\end{enumerate}

\section{Contributions}\label{contributions}

The contributions of this work are the following:

\begin{itemize}
\tightlist
\item
  Formalised a subset of Python using a methodology similar to
  \autocite{fromherz_static_2018}. The formalisation includes a subset
  of the Python syntax, an AST representation of the syntax more
  malleable to work with, and the semantics of the subset of the Python
  Language.
\item
  Defined an Abstract Domain for Variables in Python.
\item
  Defined an aliasing-aware Abstract Domain for the state of a Python
  program.
\item
  Derived semantics from the Abstract Domain, i.e.~an Abstract
  Interpreter for Python.
\item
  A working open source implementation of the Abstract Interpreter
  nicknamed Pytropos.
\end{itemize}

\section{Thesis Structure}\label{thesis-structure}

The following parts of this document are as follows:

\begin{itemize}
\tightlist
\item
  Chapter~\ref{background} explores some background material on
  Dynamically typed and Statically typed languages, Tensors and why are
  they important, the Numpy library, and Abstract Interpretation,
\item
  Chapter~\ref{pytropos-analysis-implementation} presents some of the
  gory details of the implementation of the Abstract Interpreter,
\item
  Chapter~\ref{validation-and-discussion} focuses on the tests made to
  ensure that the implementation follows the Abstract Interpreter, to
  showcase the utility of Pytropos to detect errors when using NumPy
  arrays, and to point some of the areas of improvement of the tool,
\item
  Chapter~\ref{related-work} presents some related work on Abstract
  Interpretation for Python, Static Analysis of Tensor shapes and
  related libraries Static Analysis tools, and finally,
\item
  Chapter~\ref{conclusions} sums up the work done and future directions
  to it.
\item
  Appendix~\ref{appendix-ai-theory}
  is divided into three parts: the definition of the syntax and
  semantics of a subset of Python, the definition of an Abstract
  Interpreter for Python, and finally some other ways in which
  Statically Analyse the shape of tensors for Python,
\end{itemize}
