\chapter{Related Work}\label{related-work}

A big, widely used, and mature language like Python has no lack of Static Analysis tools.
A big, widely used, and mature area like Machine Learning has no lack of people trying to
build tools to make writing code for it easier. In this chapter, a brief overview of the
many Static Analysis tools developed for Python code and the several approaches taken to
solve the problem of tensor shapes.

\section{Static Analysis in Python}

% {\inlinetodo{add note in the difficulty of knowing what is the theory
% behind some of the most widely used linters}}

In the Table~\ref{relatedworktable}, a list of all analysis tools found in the literature
and libraries is given. The usage of a tool refers to how is the tool used by developers:
embeded in the python interpreter, used as a linter, or as a library that runs with the
code to analyse. The Analysis base of a tool refers to which is the theory behind the
tool, how it works underneat.
% The purpose of analysis tools are varied and broad in scope.

\begin{sidewaystable}[p]
\centering
\begin{threeparttable}
\begin{longtable}[]{|l|l|l|l|}
  \caption{Analysis tools for Python.
  }\label{relatedworktable}\tabularnewline
  \toprule
  Tool Name / Source & Usage & Analysis base & Purpose \tabularnewline
  \midrule
  \endhead
    \textcite{cannon_localized_2005}                     & Embeded & Type analysis           & Typecheking     \tabularnewline
    Pyflakes \autocite{pyflakes2005}                     & Linter  & N/A                     & Various checks  \tabularnewline
    Pylint \autocite{thenaultpylint}                     & Linter  & N/A                     & Various checks  \tabularnewline
    Pychecker \autocite{norwitzpychecker}                & Linter  & N/A                     & Various checks  \tabularnewline
    MyPy \autocite{lehtosalo2016mypy}                    & Linter  & Gradual Type Analysis   & Type checking   \tabularnewline
    Enforce\tnote{1}\tnote{+} (defunct)                  & Library & Gradual Type Analysis   & Type checking               \tabularnewline
    Sagitta\tnote{2}\tnote{+} (defunct)                  & Library & Type analysis           & Type checking               \tabularnewline
    StyPy \autocite{ortin_towards_2015}                  & Linter  & Novel                   & Type checking               \tabularnewline
    Lyra\tnote{3}                                        & Library & Abstract Interpretation & Various analyses (including Value Analysis) \tabularnewline
    Pytropos\tnote{4} (This Work)                        & Linter  & Abstract Interpretation & Value Analysis (specialised on tensor shapes) \tabularnewline
    PyType\tnote{5}                                      & Linter  & N/A                     & Type checking and inference \tabularnewline
    ICBD\tnote{6} (defunct)                              & Linter  & N/A                     & Type checking and inference \tabularnewline
    Pyre\tnote{7}                                        & Linter  & Abstract Interpretation & Type checking               \tabularnewline
    Nagini\tnote{8} \autocite{eilers_nagini_2018}        & Linter  & SMT (Viper)             & Verifier                    \tabularnewline
    Typpete\tnote{9} \autocite{hassan_maxsmt-based_2018} & Linter  & SMT                     & Type inference              \tabularnewline
    \textcite{fromherz_static_2018}\tnote{*}             & N/A     & Abstract Interpretation & Value Analysis | Verifier   \tabularnewline
    \textcite{monat_static_2018}\tnote{*}                & N/A     & Abstract Interpretation & Type Analysis               \tabularnewline
    PyAnnotate\tnote{+}                                  & Library & Gradual Type Analysis   & Type inference              \tabularnewline
  \bottomrule
\end{longtable}
\begin{tablenotes}
  \item[+] \footnotesize It is not an static analysis. It is a dynamic analysis.
  \item[*] Tool has not been made public.
  \item[1] Homepage: \url{https://github.com/RussBaz/enforce}
  \item[2] Homepage: \url{https://github.com/peterhil/sagitta}
  \item[3] Homepage: \url{https://github.com/caterinaurban/Lyra}
  \item[4] Homepage: \url{https://github.com/helq/pytropos}
  \item[5] Homepage: \url{https://github.com/google/pytype}
  \item[6] Homepage: \url{https://github.com/kmod/icbd}
  \item[7] Homepage: \url{https://github.com/facebook/pyre-check}
  \item[8] Homepage: \url{https://github.com/marcoeilers/nagini}
  \item[9] Homepage: \url{https://github.com/caterinaurban/Typpete}
  \item[10] Homepage: \url{https://github.com/dropbox/pyannotate}
\end{tablenotes}
\end{threeparttable}
\end{sidewaystable}

\inlinetodo{write a brief description of each one of the tools presented in the table}

% {\inlinetodo{explain special case of the library \texttt{astroid} (part
% of pylint), which is able to perform some Static Value Analysis. It
% stops when it cannot determine the value of anything else. Pretty nice.
% \texttt{astroid.extract\_node(\textquotesingle{}a=1\textbackslash{}nb=2\textbackslash{}nc=a+b\textbackslash{}nc\textquotesingle{});\ next(\_.infer())}.
% https://astroid.readthedocs.io/en/latest/inference.html}}
%
% \inlinetodo{explain differences between lyra, fromherz..2018 and Pytropos}
% \inlinetodo{mention about the four papers that propose semantics for python}
%
% {\inlinetodo{lyra was not the first to try abstract interpretation for
% Python, mention pydrogen https://github.com/lapets/pydrogen}}


% - Localized Type Inference [@cannon_localized_2005]: Cannon was the first (to my
%   knowledge) to try to statically infer and type check code in Python.  His idea was to
%   infer variables’ types before executing the code, and use this knowledge to speed up
%   execution time. He acknowledges that the nature of python makes type inference to be
%   very weak and therefore the complexity that the project brought did not improve
%   performance as much as he hoped. This work is previous to @siek_gradual_2006 Gradual
%   Types. Recent work has been focused on Gradual Types, see below.
%
% - Mypy [@lehtosalo2016mypy]: it is a static typing tool, focused
%   on type checking code annotated with types. The mypy group has had an impact in Python
%   itself, since version 3.5 of python a library developed by mypy (typing) has been added
%   to the official list of Python libraries. Being mypy as relevant and impactful as it is,
%   I discuss about it deeper in the subsection below.
%
% - Enforce[^enforce]: Checks for type mismatches
%   at runtime. It checks every time that a function is called if the annotated types
%   correspond with the types of the data that is passed to the functions. It is not
%   different to adding your own asserts in code to check if the data passed is what the
%   functions expects.  Enforce may be not more than a assertion tool, but it integrates
%   with existing python type annotations and it is simpler to use than asserts.
%
% Mypy is a tool developed by the mypy group led by Jukka Lehtosalo, Guido van Rossum, and
% others. Mypy developments (and findings) have influenced the latest versions of Python. In
% the version 3.5 of the language, a new library (typing), taken from mypy has been added to
% the official library. Mypy was originally the result of the master’s thesis of Jukka
% Lehtosalo, but it has got many contributors since, one of which is Guido van Rossum, the
% benevolent dictator for life (creator and principal maintainer) of Python. One of the
% goals of mypy is to make easy for developers to type check code.
%
% However useful Mypy is, it suffers a lack of user support, which means it still remains as
% an experimental and semi-hard tool to use. Adding type annotations to Python code is often
% hard, but not because of the complexity of Mypy’s type system, rather because type
% annotations to existing code are hard to find, thus annotating any piece of code requires
% writing too the type annotations for the libraries used.
%
% Mypy has already been used in the industry, where it is fundamental to ensure the code has
% no bugs (two examples of this, are dropbox and instagram). Type annotating code in a large
% source code bases may be a monumental task, thus different companies have created tools to
% generate type annotations automatically from sample executions. Two of these tools are
% dropbox’ library pyannotate[^7], and instagram’s MonkeyType[^8].
%
% [^7]: Available at <https://github.com/dropbox/pyannotate>
% [^8]: Available at <https://github.com/Instagram/MonkeyType>
%
% The goal of type annotating code is to reduce the number of potential bugs caused by
% mismatched types. Paraphrasing Gradual Types’ principal thesis, stricter and more precise
% types should make harder for the programmer to make mistakes and should catch bugs before
% running the code. The Gradual Types’ thesis (applied to mypy) gets challenged by
% @gopinath_how_2017, who present their analysis (although with a very small code base) of
% python code. They analyze python code with type annotations vs extensive testing. As
% expected, a very good test base captures all possible bugs that the type system could
% capture.  However, @gopinath_how_2017 analysis do not make Mypy any less useful. In fact
% one may argue that the power of a type system is not only to assert that code written has
% no type mismatches, but it also helps the developer to refactor code by showing which
% parts of the code must be modified if one wants to change the type of variables and
% functions. Type annotations also serve as a way to document the inputs and outputs of a
% function or method.

\section{Tensor Shape Analysis}\label{tensor-shape-analysis}

% {\inlinetodo{add idea from http://nlp.seas.harvard.edu/NamedTensor
% Alexander, who suggests building the libraries in a different way so we
% do not have some of the problems of working with nameless shapes}}
%
% {\inlinetodo{extend solutions described below with the new libraries you
% found, comment below}}
%  {2018.06.25} tensors papers and libraries
%  * (mypy-python) https://github.com/machinalis/mypy-data/tree/master/numpy-mypy
%  * (python (in java)) ariadne https://github.com/wala/ML (paper by dolby (weird name))
%  * (haskell) https://github.com/achirkin/easytensor
%  * (nopython) https://github.com/Diderot-Language/diderot (Diderot: a domain-specific language for portable parallel scientific visualization and image analysis)
%                                                           Chiw et al (2018) - Compiling Diderot

\subsection{Solutions in other languages}%
\label{solutions-in-other-languages}

We may think, If a type system is not expressive enough to capture the
shape of tensors, then we should just start writing code in a language
which does. We may write code in a language like Haskell or even C++,
which have very strong and expressive type systems (it is possible in
C++ to enforce the shape of the tensors with the use of templates and
constraints (C++20)).

In fact, solutions in other languages exists. For example,
\textcite{chen_typesafe_2017} type checks the shape of tensors
(operations) by restricting what can be constructed (via constraints in
the types of objects and functions). Chen's solution uses the powerful
type system of Scala (which runs over the JVM).
\textcite{eaton_statically_2006} does the same, although in an old
version of Haskell. Eaton's encoding of tensor shapes is awkward because
Haskell did not have, at the time, support for natural numbers at the
type level. Haskell also lacked on syntactic sugar for type functions.
\textcite{abe_simple_2015} implement type checking for matrices (aka,
tensors) in OCaml. The library detects the shape of the matrices at
compile time. \textcite{rakic_statically_2012} type check tensor shapes
(they call them matrices) with templates in C++. Templates are only
accessible at compile time, thus the Rakić et al.~library type checks at
compile.

One recent effort to type check tensor shapes in Haskell is the library
\pycode|tensorflow-haskell-deptyped| written by
\textcite{elkin_haskell_2018}. The library is written as a wrapper
around the library port of TensorFlow for Haskell.
\pycode|tensorflow-haskell-deptyped| enforces at the type level how and
which results a operation between compatible tensors is to be performed.

A different path to take is to extend an existing type checking system,
like MyPy, and extend it with dependent types. Dependent types allow us
to carry information from the term level to the type level, i.e.~we can
encode information of our data (available only at runtime) into the type
system. Over this extended type system, we could implement some
restrictions to the operations applied to different types (this is in
fact the strategy taken in the library tensorflow-haskell-deptyped).

\subsection{Theoretical Solutions}\label{theoretical-solutions}

The problem of mismatched shapes is not new, in fact, it is so common
that it has appeared several times with similar solutions
\autocites{arnold_specifying_2010}{griffioen_type_2015}{rink_modeling_2018}{slepak_array-oriented_2014}{trojahner_dependently_2009}.
All solutions, though, are theoretical, they propose type systems which,
if were to be implemented, could warranty type safety, i.e.~no
mismatching of tensor shapes could ever happen at runtime.

The following is a small discussion about the different solutions
proposed to type check tensors found in the literature\footnote{As a
  side note, it is interesting to notice how difficult is to communicate
  ideas in science. All the papers presented in this section hope to
  solve the same (or similar) problem, but they do not reference each
  other, which means that none of them knew what the others were doing.
  The principal reason for this, I think, it is because they all called
  the problem differently.}:

\begin{itemize}
\item
  \textcite{trojahner_dependently_2009}: A paper on type checking of
  arrays. They define all type restrictions using dependent types
  (something that no other paper does). The special keyword of this work
  is \enquote{array programming}.
\item
  \textcite{arnold_specifying_2010}: A paper focused on type checking of
  sparse \enquote{matrix} operations. They define a functional language
  that can be translated into a lower level language, or machine code.
  They give a complete formalization of the algorithms and proofs of
  correctness in Isabelle. The special keyword of this work is
  \enquote{sparse matrix}.
\item
  \textcite{griffioen_type_2015}: A paper focused in type checking and
  inference of arrays in array programming and vector spaces. They
  define a special type system in which tensors are first class
  citizens. The algorithm used for type checking is
  \enquote{Unification} which allows them to infer the type shape of
  arrays. The special keyword of this work is \enquote{array
  programming}.
\item
  \textcite{slepak_array-oriented_2014}: A paper that tries to formalise
  array-oriented programming languages and extend them with unit-aware
  operations. Array-oriented programming languages are languages which
  base all their computations in arrays (like matlab), but they usually
  aren't formalised. The types of arrays not only carry information
  about their shape but also of the unit they carry. The special
  keywords of this work are \enquote{array-oriented programming} and
  \enquote{unit-aware computation}.
\item
  \textcite{rink_modeling_2018}: In this paper, Rink formalises a type
  system to type check the shape of tensors and defines a language to
  use with the type system. It is a custom type system which does not
  require of dependent types. The special keyword of this work is
  \enquote{tensor manipulation language}.
\end{itemize}
