\chapter{Conclusions and Future Work}\label{conclusions}

\section{Conclusions}

In this work, we have presented the definition and implementation of an Abstract
Interpreter for Python focused on the analysis of tensor shapes.

We showed how can Abstract Interpretation be applied to a dynamically typed programming
language like Python with the goal of defining a Static Value Analysis. To make this
possible, first, we specified small-step (concrete) semantics for Python taking into
account Python dynamically typed variables and aliasing. To build the Abstract Interpreter
we defined an apt Abstract Domain for all values in the language, an Abstract Domain for a
state of a program with aliasing, and abstract semantics. Our approach to build the
Abstract Interpreter is easily extendable to check for more than built-in values as we
showed by extending the syntax and semantics of Python to handle the shape of NumPy arrays
(tensors). The Abstract Interpreter is able to use type annotations from the user to get
more precision on the values it is able to analyse.

We presented a working implementation of the Abstract Interpreter, Pytropos, which is able
to analyse the shape of tensors in a variety of scenarios. The interpreter can work as a
linter for IDEs catching potential errors as the developer codes.

In brief, the work done includes:

\begin{itemize}
\tightlist
\item A specification of a subset of Python 3.6 (Syntax and small-step semantics).
\item An Abstract Domain for Python Values.
\item An Abstract Domain for Python program states.
\item The semantics of an Abstract Interpreter for Python.
\item The implementation of the Abstract Interpreter, Pytropos\footnote{Available at:
    \url{https://github.com/helq/pytropos}}.
\item An application for the Abstract Interpreter to statically analyse the shapes of
  tensors and tensor operations.
\item A series of tests showcasing the abilities of the Abstract Interpreter and its
  failures.
\item A way for developers to annotate code to improve the accuracy of the Abstract
  Interpreter.
\end{itemize}

The presented Python formalisation is easily extendible to calculate the value of a
library-defined class. NumPy arrays were added to the formalisation to show how to analyse
code the shape of tensors.

%The State Abstract Domain is wholy defined by the join operation.
%An Abstract Domain for the state of the program was designed. The Abstract Domain was
%built to take into account aliasing. If there is some inconsistency between two states
%between two states are collapsed into a $\top$ value. $\top$ values can be think of as
%\verb|?| types in Gradual Type systems.

The abstract semantics, the semantics of the Abstract Interpreter, are easy to develop. It
was shown how to get from the concrete semantics (small-step semantics) to the abstract
semantics.

Testing showed that the Pytropos, the implementation of the Abstract Interpreter, is able
to check many common shape mismatches. The biggest problem of the implementation
is on the lack of support for built-in functions and values.

When the Abstract Interpreter cannot compute the correct tensor shape, the user may help
the Abstract Interpreter with Type Annotations. If the user gives a type annotation that
does not improve the precision of the computed value, the interpreter will warn the user
of his mistake.

\section{Future Work}\label{future-work}

Python is a huge and rich language. The amount of characteristics that Python has exceeds
by far what a single human can implement in the span of a master thesis.

Much work is left to improve Pytropos so that it can be used by the regular developer. We
propose the following roadmap to continue building the Abstract Interpreter:

%\inlinetodo{Defining an abstract interpreter for sets}

\begin{itemize}
\tightlist
\item Extend Pytropos to include Exception handling. A similar approach to that of
  \textcite{fromherz_static_2018} could be a good starting point.
\item Improve how copying and \textbf{join}ing stores (states of the program) are done.
  The join operation between stores is very, very costly. Walking throught the graphs
  becomes prohibitely expensive as the program to analyse grows in number of \pycode|if|
  and \pycode|while| sentences. This associated cost could be reduced if only
  \enquote{diff}s of stores were used. One way to do this is by using immutable
  structures for all values in the implementation. Using immutable structures would
  require the explicit implementation of the heap and garbage collector.
\item Extend Pytropos to handle breaking control statements (\pycode|continue|,
  \pycode|break| and \pycode|return|). \textcite{fromherz_static_2018} also present a way
  to handle breaking control statement in Python.
\item Extend the global scope to handle local and nonlocal scopes. The scope rules of
  Python are mildly complex with four different variable scopes: global, local, nonlocal,
  and class. Something to take into account is the ability of CPython to statically
  analyse the use of local and non-local variables before interpreting the code.
\item Extend \(\mathbf{Val}^\sharp\) with user-defined functions and objects. Once the
  interpreter handles user-defined functions properly, extending the interpreter to work
  with user-defined objects is not a big problem. The biggest difficulty with
  objects is implementing the MRO rules in charge of how to inheritance work in Python.
\end{itemize}

% \inlinetodo{add idea that Pytropos could be run inside an IDE just as paper: Sulir Poruban
%   (2018) - Augmenting Source Code Lines with Sample Variable Values}
% \inlinetodo{add idea that the parser could be extended to add holes, or insert Top values,
%   just as done in paper: Omar et al (2018) - Live Functional Programming with Typed Holes
%   (I wrote something about it in the SoRI)}
% \inlinetodo{The Python Regression tests are very broad in scope, they use too many builtin
%   capabilities which makes it hard for a prototype like Pytropos to make any of them work
%   (and that is maybe why they are the regression tests for Python, after all they need to
%   check as many characteristics as possible!).}

Besides what is left to do to make Pytropos more powerful, there are some tasks related to
the formality of the work. This work presented some concrete and abstract semantics for
Python but there was never a proof of their correctness. We consider the following to be the
problems to solve to prove the formalisation correct:

\begin{itemize}
\tightlist
\item Give a complete and through formal definition for a subset of Python in a
  proof-assistant language such as Agda, Idris, Coq, and Isabelle/HOL.
\item Define in the same proof-assistance language the Abstract Domain, the properties it
  must follow, and the abstract semantics.
\item Prove that the abstract semantics are in fact consistent with the concrete semantics
  and abstract domain.
\end{itemize}

% Such a formalisation of Python would likely/hopefully help future
% endeavours on proofs and verification.

% Notice that any formalisation of Python must take some stance on how closely it wants to
% follow the implementation of CPython. CPython has many, many little undocumented semantic
% subtleties. Trying to write a full formal definition of Python semantics is probably
% undoable.
